Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Re-knit pdf

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
devtools::install_github("hanneoberman/miceheckman")
```

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(miceheckman)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

Load the data:
```{r obesity}
data("obesity", package = "miceheckman")
```

Select relevant variables:
```{r bmi}
dat <- obesity
```

>>>>

title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This tutorial illustrates the imputation of incomplete multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
# devtools::install_github("hanneoberman/miceheckman")
```

@@

Set up the R environment and load the necessary packages:
```{r env, message=FALSE, warning=FALSE}
set.seed(2022)        # for reproducibility
library(mice)         # for imputation
library(miceadds)     # for additional imputation routines
library(ggmice)       # for incomplete/imputed data visualization
library(ggplot2)      # for visualization
library(dplyr)        # for data wrangling
library(lme4)         # for multilevel modeling
library(mitml)        # for multilevel parameter pooling
library(micemd)  # for imputation cf. heckman models
library(metamisc)     # for case study data
```
<!-- library(broom.mixed)  # for multilevel estimates -->

@@

Load the data:
```{r obesity}
data("data_heckman", package = "micemd")
```

Select relevant variables:
```{r bmi}
dat <- data_heckman
```

### 2023-10-02 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Re-knit pdf

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\bottomrule()
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\bottomrule()
\end{longtable}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
R> devtools::install_github("hanneoberman/miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis, square = TRUE)
\end{CodeInput}
\begin{figure}

@@

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `rt`` Response time in minutes (inclusion-restriction variable).
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("obesity", package = "miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- obesity
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred["Weight","Cluster"]<- -2 # clustering variable
R> pred["Weight", ] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

@@

\begin{itemize}
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\item
  FIML vs MI
\end{itemize}

@@

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})

@@

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modelled by a random effect so
the hierarchical model can be estimated even when there are few
observations per cluster. However, this strategy is best suited for
balanced data (Grund, 2017) and when random effects model is
appropriated, i.e.~the number of clusters is adequate. (Austin,2018).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Formula \pkg{lme4}}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
\texttt{x1} predictor with random intercept \\
& varying among \texttt{g1} \\
\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
\texttt{x1} and \texttt{x2} only in fixed effect \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
of \texttt{x1} and \texttt{x2} only in fixed effect \\
& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
variance-covariance matrix estimated only with \\
& the variance terms of intercept, slope of \texttt{x1}, \\
& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
& variance-covariance matrix estimated separately, \\
& i.e, one for intercept and \texttt{x1} and another for \\
& intercept and \texttt{x2} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with correlated random intercept and \\
& random slope of \texttt{x} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with uncorrelated random intercept \\
& and random slope of \texttt{x1} \\
\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
& Random intercept varying among \texttt{g1} and among
\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
& \\
\bottomrule()
\end{longtable}

>>>>

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\end{longtable}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
R> # devtools::install_github("hanneoberman/miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(micemd)  # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis)
\end{CodeInput}
\begin{figure}

@@

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

@@

The matrix only shows the predictors for the main model, not the
selection model.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred[,"Cluster"]<- -2 # clustering variable
R> pred[, "Weight"] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

@@

ORDER:

\begin{itemize}
\item
  summary
\item
  congeniality, then in hierarchical models
\item
  look whether we can fit cong. back in the main body
\item
  alt. methods
\item
  conclusion: mice is really easy!
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\item
  FIML vs MI
\end{itemize}

@@

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
\url{https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/})
TODO: replace ref.

@@

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modeled by a random effect so the
hierarchical model can be estimated even when there are few observations
per cluster. However, this strategy is best suited for balanced data
(Grund, 2017) and when random effects model is appropriated, i.e.~the
number of clusters is adequate. (Austin,2018).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.5495}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.4505}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Formula \pkg{lme4}}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\texttt{y\ \textasciitilde{}\ x1\ +\ (1\ \textbar{}\ g1)} & Fixed
\texttt{x1} predictor with random intercept \\
& varying among \texttt{g1} \\
\texttt{y\textasciitilde{}x1*x2+\ (1\textbar{}\ g1)} & Interactions of
\texttt{x1} and \texttt{x2} only in fixed effect \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x2\textbar{}\ g1)} & Interactions
of \texttt{x1} and \texttt{x2} only in fixed effect \\
& with slope of \texttt{x2} randomly varying among \texttt{g1} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1*x2\textbar{}\ g1)} &
variance-covariance matrix estimated only with \\
& the variance terms of intercept, slope of \texttt{x1}, \\
& slope of \texttt{x2} and interaction \texttt{x1*x2} \\
\texttt{y\ \textasciitilde{}\ x1*x2+\ (x1\ \textbar{}\ g1)+\ (x2\textbar{}\ g1)}
& variance-covariance matrix estimated separately, \\
& i.e, one for intercept and \texttt{x1} and another for \\
& intercept and \texttt{x2} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with correlated random intercept and \\
& random slope of \texttt{x} \\
\texttt{y\ \textasciitilde{}\ x1\ +\ (x1\ \textbar{}\textbar{}\ g1)\ or\ 1\ +\ x1\ +\ (1\ \textbar{}\ g1)\ +\ (0\ +\ x1\ \textbar{}\ g1)}
& Fixed \texttt{x1} with uncorrelated random intercept \\
& and random slope of \texttt{x1} \\
\texttt{y\ \textasciitilde{}\ (1\ \textbar{}\ g1)\ +\ (1\ \textbar{}\ g2)}
& Random intercept varying among \texttt{g1} and among
\texttt{g2\ \ \ \textbar{}\ \textbar{}}y \textasciitilde{} (1 \\
& \\
\end{longtable}
